Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-15 paper/paper.Rmd
moved everything to R Mardkown and now template works out well with rticles

---
title: "Overlooked dynamics of algorithmic recourse"
author:
- "Patrick Altmeyer (p.altmeyer@tudelft.nl)"
date: "`r format(Sys.Date(), '%B, %Y')`"
output: 
  bookdown::pdf_document2: 
    toc: false
    includes:
      in_header: ../tex/preamble.tex
      before_body: ../tex/before_body.tex
  bookdown::html_document2: default
fontsize: 12pt
bibliography: ../bib.bib
link-citations: true
abstract: "..."
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  out.width = 450,
  fig.align = "center"
)
```

```{r introduction, child="introduction.Rmd"}
```

\FloatBarrier
\pagebreak

# References {-}

<div id="refs"></div>

\pagebreak

```{r annex, child="appendix.Rmd"}
```

>>>>

---
title: Endogenous Macrodynamics in Algorithmic Recourse
affiliation:
  ## use one only of the following
  # author-columnar: true         ## one column per author
  institution-columnar: true  ## one column per institution (multiple autors eventually)
  # wide: true                  ## one column wide author/affiliation fields

  #author:   ## A custom author ordering field that breaks institution grouping.
  #  - name: Eldon Tyrell
  #    mark: 4
  #    email: eldon@starfleet-academy.star
  #  - name: Michael Shell
  #    mark: 1
  #  - name: Roy Batty
  #    mark: 4
  #    email: roy@replicant.offworld
  ## Authors cited in institution field are ignored when author field exists

  #author_multiline: ## Put authors in a given order, with multiline possibility. Authors cited in institution are ignored if exists
  #  - line:         ## Create a new author line
  #    - name: Michael Shell
  #      mark: 1
  #    - name: Homer Simpson
  #      mark: 2
  #      email: homer@thesimsons.com
  #    - name: Roy Batty
  #      mark: 4
  #      email: roy@replicant.offworld
  #  - line:         ## Create a second author line
  #    - name: Montgomery Scott
  #      mark: 3
  #    - name: Eldon Tyrell
  #      mark: 4

  # institution:
  #   - name: Georgia Institute of Technology
  #     department: School of Electrical and Computer Engineering
  #     location: Atlanta, Georgia 30332--0250
  #     email: ece@datech.edu
  #     mark: 1
  #     author:
  #       - name: Michael Shell
  #   - name: Twentieth Century Fox
  #     location: Springfield, USA
  #     mark: 2
  #     author:
  #       - name: Homer Simpson
  #         email: homer@thesimpsons.com
  #   - name: Starfleet Academy
  #     location: San Francisco, California 96678-2391
  #     other: "Telephone: (800) 555--1212, Fax: (888) 555--1212"
  #     mark: 3
  #     author:
  #       - name: Montgomery Scott
  #   - name: Tyrell Inc.
  #     location: 123 Replicant Street, Los Angeles, USA 90210--4321
  #     mark: 4
  #     author:
  #       - name: Eldon Tyrell
  #         email: eldon@starfleet-academy.star
  #       - name: Roy Batty
  #         email: roy@replicant.offworld
keywords: ["keyword 1", "keyword 2"]
abstract: |
  Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model the goal is to find valid counterfactuals for individual instance that fulfill various desiderata. The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point. There has also been surprisingly little work on the related question of how the actual implementation of recourse by one individual may affect other individuals. Through this work we aim to close that gap by systematizing and extending existing knowledge. We first show that many of the existing methodologies can be collectively described by a generalized framework. We then argue that the existing framework fails to account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level. Through simulation experiments involving various popular counterfactual generators and several benchmark datasets, we generate a total XX million Counterfactual Explanations and study the resulting domain and model shifts. We find that the induced shifts are substantial enough to likely impede the applicability Algorithmic Recourse in practice. Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches. Our simulation framework for studying recourse dynamics is fast and open-sourced.

# use some specific Tex packages if needed. 
# with_ifpdf: true
# with_cite: true
# amsmath need to be true to use with bookdown for referencing equations.
with_amsmath: true
# with_algorithmic: true
# with_array: true
# with_dblfloatfix: true

bibliography: https://raw.githubusercontent.com/pat-alt/bib/main/bib.bib
output: 
  bookdown::pdf_book:
    base_format: rticles::ieee_article
    includes:
      in_header: "tex/preamble.tex"
      before_body: "tex/before_body.tex"
#citation_sorting: none   ## used as sorting option of the biblatex package (if selected)
---

```{r, include=FALSE}
child_docs <- c(
  "sections/introduction.rmd",
  "sections/related.rmd",
  "sections/methodology.rmd",
  "sections/methodology_2.rmd",
  "sections/empirical.rmd",
  "sections/empirical_2.rmd",
  "sections/mitigation.rmd",
  "sections/discussion.rmd",
  "sections/limitations.rmd",
  "sections/conclusion.rmd"
)
knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  fig.align = "center",
  out.width = "90%"
)
```

```{r, child=child_docs}
```

# References {.unnumbered}

::: {#refs}
:::

{{< include sections/appendix.qmd >}}
